---
title: "Microsim Development of Person Initialization Models"
output: html_document
date: "2024-06-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(mlogit)
library(ordinal)
library(MASS)
library(brant)
setwd("/Users/deligkaris.1/OneDrive - The Ohio State University Wexner Medical Center/MICROSIM/NOTEBOOKS/DATA")
df <- read.csv("nhanesDf.csv",header=TRUE)
```
## Goal

The goal of this notebook is to develop models that can be used by Microsim

### NHANES dataframe

The NHANES dataframe has been loaded to memory and we now need to specify which columns are categorical (unordered by default) and
which ones are categorical and ordered.
In addition, during the development of the waist model, I noticed that waist was dependent on a non-linear bmi term so I am
calculating bmi*bmi here as well.

```{r}
df$education <- ordered(df$education)
df$gender <- factor(df$gender)
df$smokingStatus <- factor(df$smokingStatus)
df$anyPhysicalActivity <- factor(df$anyPhysicalActivity)
df$alcoholPerWeek <- ordered(df$alcoholPerWeek)
df$statin <- factor(df$statin)
df$raceEthnicity <- factor(df$raceEthnicity)
#bmi2 is bmi^2
df$bmi2 <- df$bmi*df$bmi

#print(levels(df$alcoholPerWeek))
names(df)

```


### Pre-processing

Let's set the year and sample from the NHANES dataframe.

```{r}
dfAllYears <- data.frame()
popSizePerYear <- 50000
yearI <- 1999
yearF <- 2017
for (iYear in yearI:yearF) {
  dfYear <- df[df$year==iYear,]
  if (nrow(dfYear)>0) {
    dfYear <- dfYear[sample(1:nrow(dfYear), popSizePerYear, replace=TRUE, prob=dfYear$WTINT2YR), ]
    dfAllYears <- rbind(dfAllYears, dfYear)
  }
} 
df <- dfAllYears
```

### Waist model

NHANES has weights included for the sample. Let's get a feeling for how much difference sampling makes in terms of
the waist distribution.
Means and standard deviations.

```{r}
#c(mean(df$waist), mean(sample(df$waist, popSize, replace=TRUE, prob=df$WTINT2YR)))
#c(sd(df$waist), sd(sample(df$waist, popSize, replace=TRUE, prob=df$WTINT2YR)))
dim(df)
```
Waist distribution.

```{r}
ggplot(df, aes(x = waist, y=after_stat(density))) + geom_histogram() +
  ggtitle("Waist distribution with sampling from NHANES 1999-2017")
```
Plot dependence of waist as a function of each continuous covariate.
```{r}
for (cov in c("age","sbp","dbp", "a1c", "hdl", "ldl", "trig", "totChol", "bmi", "bmi2", "creatinine")) {
  plot<-ggplot(df[0:5000,], aes(y=waist, x=get(cov), color=WTINT2YR)) + 
    geom_point() +
    #xlim(50, 250) +
    #ylim(50, 250) +
    ggtitle(paste0("Waist in NHANES vs ", cov))
  print(plot)
}
```
Let's fit a linear model using the NHANES data.

```{r}
waistModel <- lm(waist ~ raceEthnicity + education + gender + smokingStatus + age + sbp + dbp + a1c + hdl + 
                   ldl + trig + totChol + bmi + anyPhysicalActivity + 
                   alcoholPerWeek + creatinine + statin + antiHypertensiveCount, 
                 data=df,)
                 #weights=df$WTINT2YR)
summary(waistModel)
```
Results showed three dominant factors: gender, age, bmi.
Let's fit a linear model using only those variables.
By including only a linear bmi term, the residuals clearly show to follow a trend and the addition of the bmi*bmi term
improves the results significantly.
I tried incorporating interactions but nothing else was found to make a significant difference.

```{r}
waistModel <- lm(waist ~ gender + age + bmi + bmi2, 
                 data=df, )
                 #weights=df$WTINT2YR)
summary(waistModel)
```
The model predicts waist fairly well.
One approach for taking care of the overstimations at large waists is to set a cap in the simulation (which we already do I think).

```{r}
waistMax <- max(df$waist)
df$waistPredicted <- predict(waistModel, data=df)
#df$waistPredicted <- pmin(df$waistPredicted, waistMax)
ggplot(df, aes(x=waist, y=waistPredicted, color=WTINT2YR)) + 
  geom_point() +
  xlim(50, 250) +
  ylim(50, 250) +
  ggtitle("Predicted waist vs waist in NHANES")
```
Residuals seem to be well normally distributed.
```{r}
df$waistResiduals <- waistModel$residuals
#df$waistResiduals <- df$waist - df$waistPredicted
ggplot(df, aes(x = waistResiduals, y=after_stat(density))) + geom_histogram() +
  ggtitle("Waist residual distribution.")
```
There is a small trend evident in this plot of residuals, much improved though compared to the model without the non-linear bmi term.
```{r}
df$waistResiduals <- waistModel$residuals
#df$waistResiduals <- df$waist - df$waistPredicted
ggplot(df, aes(x = waistPredicted, y=waistResiduals, color=bmi)) + geom_point()
  ggtitle("Waist residual distribution.")
```
Add to our expected waist prediction a draw from the normal distribution 
N(mean of residuals, sd of residuals).
Plot the three distributions on the same plot.

```{r}
nPoints <- nrow(df)
residualsMean <- mean(waistModel$residuals)
residualsSd <- summary(waistModel)$sigma
draws <- rnorm(n=nPoints, mean=residualsMean, sd=residualsSd)
df$draws <- draws
df$waistPredictedWithDraws <- df$waistPredicted + df$draws
dfLong <- data.frame(values = c(df$waist, df$waistPredicted, df$waistPredictedWithDraws), 
                     group = c(rep("waist", nPoints), rep("waistPredicted", nPoints), rep("waistPredictedWithDraws", nPoints)))

ggplot(dfLong, aes(x = values, y=after_stat(density), fill = group)) + 
  geom_histogram(alpha = 0.5) +
  xlim(50, 150)
```
## Education model

```{r}
for (cov in c("sbp", "dbp", "age",  "a1c", "hdl", "ldl",  "trig", "totChol", "bmi", "waist", "creatinine" )) {
  plot<-ggplot(df, aes(x=education, y=get(cov))) + 
    geom_violin() + 
    geom_boxplot(width=0.1, fill="white") +
    ylab(cov)
    #ggtitle(paste0())
  print(plot)
}
```

```{r}
for (cov in c("raceEthnicity", "gender", "smokingStatus", "anyPhysicalActivity", "alcoholPerWeek", "statin", "antiHypertensiveCount")) {
  plot<-ggplot(df, aes(x=education, fill=get(cov))) + geom_bar(position="dodge") 
  print(plot)
}
```


Fit an ordered logistic model for education.

```{r}
educationModel <- polr(education ~ waist + raceEthnicity + gender + smokingStatus + age + 
                           sbp + dbp + a1c + hdl + ldl + trig + totChol + bmi + anyPhysicalActivity + 
                           alcoholPerWeek + creatinine + statin + antiHypertensiveCount, data=df, Hess=TRUE)
summary(educationModel)
```
The dominant factors seem to be race/ethnicity, smokingStatus, and sbp or age or alcoholperweek.
```{r}
educationModel <- polr(education ~ raceEthnicity + smokingStatus + age , data=df, Hess=TRUE, method="logistic")
summary(educationModel)
```

```{r}
brant(educationModel)
```
```{r}
#test <- predict(educationModel, df, type="p")
df$educationPredicted <- ordered(predict(educationModel, df, type="class"))

isCorrectEducation <- function(row) {
  education <- row["education"]
  educationPredicted <- row["educationPredicted"]
  if ( education == educationPredicted) {
    return(1)
  }
  else {
    return(0)
  }
}

df$correctEducation <- sapply(1:nrow(df), function(i) isCorrectEducation(df[i,]))
#df$educationPredicted <- as.numeric(df$educationPredicted) + 1
df$educationDifference <- as.numeric(df$education) - as.numeric(df$educationPredicted)
percentWithin1 <- (nrow(df[ abs(df$educationDifference)==1 ,]) + nrow( df[ df$educationDifference ==0, ]) ) / nrow(df)
c(mean(df$correctEducation), mean(df$educationDifference), percentWithin1)
```

```{r}
ggplot(df, aes(x = educationDifference, y=after_stat(density))) + geom_histogram() +
  ggtitle("Education level difference distribution.")
```
## Alcohol model

```{r}
alcoholModel <- polr(alcoholPerWeek ~ waist + education + raceEthnicity + gender + smokingStatus + age + 
                           sbp + dbp + a1c + hdl + ldl + trig + totChol + bmi + anyPhysicalActivity + 
                            creatinine + statin + antiHypertensiveCount, data=df, Hess=TRUE)
summary(alcoholModel)
```
```{r}
df$alcoholPredicted <- ordered(predict(alcoholModel, df, type="class"))

isCorrectAlcohol <- function(row) {
  alcohol <- row["alcoholPerWeek"]
  alcoholPredicted <- row["alcoholPredicted"]
  if ( alcohol == alcoholPredicted) {
    return(1)
  }
  else {
    return(0)
  }
}

df$correctAlcohol <- sapply(1:nrow(df), function(i) isCorrectAlcohol(df[i,]))
df$alcoholDifference <- as.numeric(df$alcoholPerWeek) - as.numeric(df$alcoholPredicted)
percentWithin1 <- (nrow(df[ abs(df$alcoholDifference)==1 ,]) + nrow( df[ df$alcoholDifference ==0, ]) ) / nrow(df)
c(mean(df$correctAlcohol), mean(df$alcoholDifference), percentWithin1)
```

```{r}
alcoholModel <- polr(alcoholPerWeek ~ gender + smokingStatus + age, data=df, Hess=TRUE)
summary(alcoholModel)
```

```{r}
brant(alcoholModel)
```

```{r}
df$alcoholPredicted <- ordered(predict(alcoholModel, df, type="class"))

isCorrectAlcohol <- function(row) {
  alcohol <- row["alcoholPerWeek"]
  alcoholPredicted <- row["alcoholPredicted"]
  if ( alcohol == alcoholPredicted) {
    return(1)
  }
  else {
    return(0)
  }
}

df$correctAlcohol <- sapply(1:nrow(df), function(i) isCorrectAlcohol(df[i,]))
df$alcoholDifference <- as.numeric(df$alcoholPerWeek) - as.numeric(df$alcoholPredicted)
percentWithin1 <- (nrow(df[ abs(df$alcoholDifference)==1 ,]) + nrow( df[ df$alcoholDifference ==0, ]) ) / nrow(df)
c(mean(df$correctAlcohol), mean(df$alcoholDifference), percentWithin1)
```
```{r}
ggplot(df, aes(x = alcoholDifference, y=after_stat(density))) + geom_histogram() +
  ggtitle("Alcohol level difference distribution.")
```
Now, I realized that the data in the original NHANES df are more continuous than categorical so I want to try model
that alcohol data using a lm model and not an ordered logistic model.

Alcohol continuous data distribution.

```{r}
ggplot(df, aes(x = alcoholPerWeekCon, y=after_stat(density))) + geom_histogram() +
  ggtitle("AlcoholPerWeekCon distribution with sampling from NHANES 1999-2017")
```
Fit a lm model to all covariates.

```{r}
alcoholConModel <- lm(alcoholPerWeekCon ~ raceEthnicity + education + gender + smokingStatus + age + sbp + dbp + a1c + hdl + 
                   ldl + trig + totChol + bmi + anyPhysicalActivity + 
                   waist + creatinine + statin + antiHypertensiveCount, 
                 data=df,)
                 #weights=df$WTINT2YR)
summary(alcoholConModel)
```
Fit a constrained model.
```{r}
alcoholConModel <- lm(alcoholPerWeekCon ~ raceEthnicity + education + gender + smokingStatus + age, 
                 data=df,)
                 #weights=df$WTINT2YR)
summary(alcoholConModel)
```
Visualize predictions against training data. Agreement is terrible.
But maybe once I categorize these predictions agreement will be better. But then my ordered logistic model is working fine.
```{r}
df$alcoholPerWeekConPredicted <- predict(alcoholConModel, data=df)
ggplot(df, aes(x=alcoholPerWeekCon, y=alcoholPerWeekConPredicted, color=WTINT2YR)) + 
  geom_point() +
  ggtitle("Predicted alcoholPerWeek vs alcoholPerWeek in NHANES")
```

```{r}
df$alcoholPerWeekConResiduals <- alcoholConModel$residuals
ggplot(df, aes(x = alcoholPerWeekConResiduals, y=after_stat(density))) + geom_histogram() +
  ggtitle("AlcoholPerWeek residual distribution.")
```