---
title: "Kaiser Stroke and Dementia Validation"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE, echo=FALSE, warning=FALSE, results='hide'}
knitr::opts_knit$set(root.dir = "/Users/deligkaris.1/OneDrive - The Ohio State University Wexner Medical Center/MICROSIM/NOTEBOOKS/DATA")
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(survival)
library(survminer)
library(RColorBrewer)
library(dplyr)
library(stringr)
library(tidyr)
library(knitr)
library(kableExtra)
```

## Goal

This notebook performs analyses of the simulated Kaiser population, for several outcomes.
The results from the simulation are compared with the relevant publications (Kent2021 and Kent2022).
Note that the results from Kent2021 and Kent2022 were manually extracted from the published figures.

## Stroke Incidence Rates (after 4 years of simulation)

```{r}
stroke4YrRates <- read.csv("kaiserStrokeValidation4YrRates.csv", header=TRUE)
kable(stroke4YrRates)
```

## Stroke-free Probabilities (after 13 years of simulation)


```{r}
dfStrokeTimes <- read.csv("kaiserStrokeValidation13YrTimes.csv",header=TRUE)
survival_object <- Surv(time = dfStrokeTimes$time, event = dfStrokeTimes$event)
fitsbiwmh <- survfit(survival_object ~ sbiwmhGroup, data=dfStrokeTimes)
dfStrokeProb <- data.frame(time = summary(fitsbiwmh)$time, surv = summary(fitsbiwmh)$surv, strata = summary(fitsbiwmh)$strata)
dfStrokeProb$sbiwmh <- dfStrokeProb$strata
dfStrokeProb <- dfStrokeProb |> 
          mutate(
            sbiwmh = str_replace_all(sbiwmh, "sbiwmhGroup=0", "no SBI or WMH"),
            sbiwmh = str_replace_all(sbiwmh, "sbiwmhGroup=1", "only SBI"),
            sbiwmh = str_replace_all(sbiwmh, "sbiwmhGroup=2", "only WMH"),
            sbiwmh = str_replace_all(sbiwmh, "sbiwmhGroup=3", "SBI and WMH"),
          ) |>
        mutate(probability = surv) |>
        select(time, probability, sbiwmh)
dfStrokeProb$group <- rep("Microsim", nrow(dfStrokeProb))

dfStrokeProbRef <- read.csv("kaiserStrokeValidation13YrProbSbiWmhRef.csv",header=TRUE)
dfStrokeProbRef <- dfStrokeProbRef |> 
                        pivot_longer(cols=!time, names_to="sbiwmh", values_to="probability") |>
                        mutate(
                          sbiwmh = str_replace_all(sbiwmh, "sbiNoWmhYes", "only WMH"),
                          sbiwmh = str_replace_all(sbiwmh, "sbiYesWmhNo", "only SBI"),
                          sbiwmh = str_replace_all(sbiwmh, "sbiYesWmhYes", "SBI and WMH"),
                          sbiwmh = str_replace_all(sbiwmh, "sbiNoWmhNo", "no SBI or WMH"),
                        )
dfStrokeProbRef$group <- rep("Reference", nrow(dfStrokeProbRef))
dfStrokeProbRef <- dfStrokeProbRef |> select(time, probability, sbiwmh, group)

dfStrokeProbAll <- rbind(dfStrokeProb, dfStrokeProbRef)

ggplot(data=dfStrokeProbAll) +
  aes(x=factor(time), y=probability, color=group, shape=sbiwmh) +
  #facet_wrap(~sbi, nrow=2, ncol=1) +
  geom_point() +
  scale_x_discrete(name="years") +
  scale_y_continuous(name="stroke-free probability", limits=c(0,1)) +
  theme_bw()

```

```{r}
fitSeverity <- survfit(survival_object ~ severityGroup, data=dfStrokeTimes)

dfStrokeProb <- data.frame(time = summary(fitSeverity)$time, surv = summary(fitSeverity)$surv, strata = summary(fitSeverity)$strata)
dfStrokeProb$severity <- as.character(dfStrokeProb$strata)
dfStrokeProb <- dfStrokeProb |> 
          mutate(
            severity = str_replace_all(severity, "severityGroup=0", "CTNo"),
            severity = str_replace_all(severity, "severityGroup=1", "CTNonspecified"),
            severity = str_replace_all(severity, "severityGroup=2", "CTMild"),
            severity = str_replace_all(severity, "severityGroup=3", "CTModerate"),
            severity = str_replace_all(severity, "severityGroup=4", "CTSevere"),
            severity = str_replace_all(severity, "severityGroup=5", "MRNo"),
            severity = str_replace_all(severity, "severityGroup=6", "MRNonspecified"),
            severity = str_replace_all(severity, "severityGroup=7", "MRMild"),
            severity = str_replace_all(severity, "severityGroup=8", "MRModerate"),
            severity = str_replace_all(severity, "severityGroup=9", "MRSevere"),
          ) |>
        mutate(probability = surv) |>
        select(time, probability, severity)
dfStrokeProb$source <- rep("Microsim", nrow(dfStrokeProb))

dfStrokeProbRef <- read.csv("kaiserStrokeValidation13YrProbWmhSeverityRef.csv",header=TRUE)
dfStrokeProbRef <- rename(dfStrokeProbRef, 
                          CTNo=ctNo, CTNonspecified=ctUnknown, CTMild=ctMild, CTModerate=ctModerate, CTSevere=ctSevere,
                          MRNo=mrNo, MRNonspecified=mrUnknown, MRMild=mrMild, MRModerate=mrModerate, MRSevere=mrSevere)
dfStrokeProbRef <- dfStrokeProbRef |> pivot_longer(cols=!time, names_to="severity", values_to="probability")
dfStrokeProbRef$source <- rep("Reference", nrow(dfStrokeProbRef))
dfStrokeProbRef <- dfStrokeProbRef |> select(time, probability, severity, source)

dfStrokeProbAll <- rbind(dfStrokeProb, dfStrokeProbRef)

ggplot(data=dfStrokeProbAll) +
  aes(x=factor(time), y=probability, color=source) +
  facet_wrap(~severity, nrow=3, ncol=4) +
  geom_point() +
  scale_x_discrete(name="years") +
  scale_y_continuous(name="stroke-free probability") +
  theme_bw()
```

## Dementia Incidence Rates (after 4 years of simulation)

Simulation dementia incidence rates after 4 years of simulation are compared with the published rates.
The published confidence intervals are also shown.
Note that the follow up intervals varied for each individual in the publication whereas in the simulation everyone had
exactly 4 years of outcomes.
Also, these rates were used during the optimization of the baseline risk for the dementia outcome.

```{r}
dementia4YrRates <- read.csv("kaiserDementiaValidation4YrRates.csv",header=TRUE)
kable(dementia4YrRates)
```

## Dementia-free probabilities (after 13 years of simulation)

### Stratified by SBI

Now we will look at the dementia-free probability for two groups, one with SBI and one without SBI, for a period of 13 years.
These simulation results are in good agreement with Figure 1 in Kent2022.
Note that the Reference data points were manually extracted from Figure 1 Kent2022.

```{r}
dfDementiaTimes <- read.csv("kaiserDementiaValidation13YrTimes.csv",header=TRUE)
survival_object <- Surv(time = dfDementiaTimes$time, event = dfDementiaTimes$event)
fitSbi <- survfit(survival_object ~ sbi, data=dfDementiaTimes)

dfDementiaProb <- data.frame(time = summary(fitSbi)$time, surv = summary(fitSbi)$surv, strata = summary(fitSbi)$strata)
dfDementiaProb$sbi <- dfDementiaProb$strata
dfDementiaProb <- dfDementiaProb |> 
          mutate(
            sbi = str_replace_all(sbi, "sbi=0", "without SBI"),
            sbi = str_replace_all(sbi, "sbi=1", "with SBI"),
          ) |>
        mutate(probability = surv) |>
        select(time, probability, sbi)
dfDementiaProb$group <- rep("Microsim", nrow(dfDementiaProb))

dfDementiaProbRef <- read.csv("kaiserDementiaValidation13YrProbSbiRef.csv",header=TRUE)
dfDementiaProbRef <- dfDementiaProbRef |> 
                        pivot_longer(cols=!time, names_to="sbi", values_to="probability") |>
                        mutate(
                          sbi = str_replace_all(sbi, "sbiNo", "without SBI"),
                          sbi = str_replace_all(sbi, "sbiYes", "with SBI")
                        )
dfDementiaProbRef$group <- rep("Reference", nrow(dfDementiaProbRef))
dfDementiaProbRef <- dfDementiaProbRef |> select(time, probability, sbi, group)

dfDementiaProbAll <- rbind(dfDementiaProb, dfDementiaProbRef)

ggplot(data=dfDementiaProbAll) +
  aes(x=factor(time), y=probability, color=group, shape=sbi) +
  #facet_wrap(~sbi, nrow=2, ncol=1) +
  geom_point() +
  scale_x_discrete(name="years") +
  scale_y_continuous(name="dementia-free probability", limits=c(0,1)) +
  theme_bw()
```

### Stratified by WMH

Now we will look at the dementia-free probabilities for the Kaiser subgroups with and without WMH.

```{r}
fitWmh <- survfit(survival_object ~ wmh, data=dfDementiaTimes)

dfDementiaProb <- data.frame(time = summary(fitWmh)$time, surv = summary(fitWmh)$surv, strata = summary(fitWmh)$strata)
dfDementiaProb$wmh <- dfDementiaProb$strata
dfDementiaProb <- dfDementiaProb |> 
          mutate(
            wmh = str_replace_all(wmh, "wmh=0", "without WMH"),
            wmh = str_replace_all(wmh, "wmh=1", "with WMH"),
          ) |>
        mutate(probability = surv) |>
        select(time, probability, wmh)
dfDementiaProb$source <- rep("Microsim", nrow(dfDementiaProb))

dfDementiaProbRef <- read.csv("kaiserDementiaValidation13YrProbWmhRef.csv",header=TRUE)
dfDementiaProbRef <- dfDementiaProbRef |> 
                        pivot_longer(cols=-c(time), names_to="wmh", values_to="probability") |>
                        mutate(
                          wmh = str_replace_all(wmh, "wmhNo", "without WMH"),
                          wmh = str_replace_all(wmh, "wmhYes", "with WMH")
                        )
dfDementiaProbRef$source <- rep("Reference", nrow(dfDementiaProbRef))
dfDementiaProbRef <- dfDementiaProbRef |> select(time, probability, wmh, source)

dfDementiaProbAll <- rbind(dfDementiaProb, dfDementiaProbRef)

ggplot(data=dfDementiaProbAll) +
  aes(x=factor(time), y=probability, color=source, shape=wmh) +
  #facet_wrap(~sbi, nrow=2, ncol=1) +
  geom_point() +
  scale_x_discrete(name="years") +
  scale_y_continuous(name="dementia-free probability", limits=c(0,1)) +
  theme_bw()

```

### Stratified by Modality and Severity

Overall, the simulation reproduces qualitatively at least the results shown in Figure 2 in Kent2023.

```{r}
#setwd("/Users/deligkaris.1/OneDrive - The Ohio State University Wexner Medical Center/MICROSIM/NOTEBOOKS/DATA")
#df <- read.csv("kaiserDementiaValidation.csv",header=TRUE)
#blues <- brewer.pal(4, "Blues")
#oranges <- brewer.pal(4, "Oranges")
#greens <- brewer.pal(5, "Greens")
#ct <- append(blues, values=greens[3], after=1)
#mri <- append(oranges, values=greens[5], after=1)
#palette <- c(ct, mri)
#survival_object <- Surv(time = df$time, event = df$event)
fitScd <- survfit(survival_object ~ severityGroup, data=dfDementiaTimes)

#fitLabels <- c("CT NO ", "CT UNKNOWN", "CT MILD", "CT MODERATE", "CT SEVERE",
#                         "MRI NO ", "MRI UNKNOWN", "MRI MILD", "MRI MODERATE", "MRI SEVERE")
#ggsurvplot(fit, 
#           xlab="time [years]", ylab="dementia-free probability", 
#           legend.labs=fitLabels,
#           legend=c(0.15, 0.38),
#           palette=palette,
#           break.time.by=1,
#           break.y.by=0.1,
#           ylim=c(0.4,1))$plot 

dfDementiaProb <- data.frame(time = summary(fitScd)$time, surv = summary(fitScd)$surv, strata = summary(fitScd)$strata)
dfDementiaProb$severity <- as.character(dfDementiaProb$strata)
dfDementiaProb <- dfDementiaProb |> 
          mutate(
            severity = str_replace_all(severity, "severityGroup=0", "CTNo"),
            severity = str_replace_all(severity, "severityGroup=1", "CTNonspecified"),
            severity = str_replace_all(severity, "severityGroup=2", "CTMild"),
            severity = str_replace_all(severity, "severityGroup=3", "CTModerate"),
            severity = str_replace_all(severity, "severityGroup=4", "CTSevere"),
            severity = str_replace_all(severity, "severityGroup=5", "MRNo"),
            severity = str_replace_all(severity, "severityGroup=6", "MRNonspecified"),
            severity = str_replace_all(severity, "severityGroup=7", "MRMild"),
            severity = str_replace_all(severity, "severityGroup=8", "MRModerate"),
            severity = str_replace_all(severity, "severityGroup=9", "MRSevere"),
          ) |>
        mutate(probability = surv) |>
        select(time, probability, severity)
dfDementiaProb$source <- rep("Microsim", nrow(dfDementiaProb))

dfDementiaProbRef <- read.csv("kaiserDementiaReference.csv",header=FALSE)
colNames <- dfDementiaProbRef$V1
dfDementiaProbRef <- data.frame(dfDementiaProbRef |> select("V2":"V11") |> t())
names(dfDementiaProbRef) <- colNames
dfDementiaProbRef$x <- dfDementiaProbRef$x / 12.
dfDementiaProbRef <- dfDementiaProbRef |> pivot_longer(cols=!x, names_to="severity", values_to="probability")
dfDementiaProbRef <- data.frame(dfDementiaProbRef)
names(dfDementiaProbRef)[1] <- "time"
dfDementiaProbRef$source <- rep("Reference", nrow(dfDementiaProbRef))
dfDementiaProbRef <- dfDementiaProbRef |> select(time, probability, severity, source)

dfDementiaProbAll <- rbind(dfDementiaProb, dfDementiaProbRef)

ggplot(data=dfDementiaProbAll) +
  aes(x=factor(time), y=probability, color=source) +
  facet_wrap(~severity, nrow=3, ncol=4) +
  geom_point() +
  scale_x_discrete(name="years") +
  scale_y_continuous(name="dementia-free probability") +
  theme_bw()

```

## MI Incidence Rates (after 4 years of simulation)

```{r}
mi4YrRates <- read.csv("kaiserMiValidation4YrRates.csv", header=TRUE)
kable(mi4YrRates)
```

## Death Incidence Rates (after 4 years of simulation)

```{r}
death4YrRates <- read.csv("kaiserDeathValidation4YrRates.csv", header=TRUE)
kable(death4YrRates)
```

## Linear Models for the mortality hazard ratios

Now I will find the straight lines I will use to model the time dependent hazard ratios for mortality.
These will be used in the non CV death model of microsim.

```{r}
ages <- c(56, 65, 72)
hrSbi <- c(1.34, 1.27, 1.22)
hrWmhCt <- c(1.41, 1.34, 1.28)
hrWmhMri <- c(1.26, 1.20, 1.15)
lmSbi <- lm( hrSbi ~ ages)
summary(lmSbi)
#predict(lmSbi, data.frame(ages))
lmWmhCt <- lm( hrWmhCt ~ ages)
summary(lmWmhCt)

lmWmhMri <- lm( hrWmhMri ~ ages)
summary(lmWmhMri)
```

