---
title: "Kaiser Stroke and Dementia Validation"
output:
  html_document: default
  pdf_document: default
date: "2024-08-09"
---

```{r, include=FALSE, echo=FALSE, warning=FALSE, results='hide'}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(survival)
library(survminer)
library(RColorBrewer)
```

## Goal

This notebook performs the survival analysis of the simulated Kaiser population.
The results from the simulation are compared with the relevant publications (Kent2021 and Kent2023).

## Stroke-free probability   

The stroke-free probability as a function of time in the simulated Kaiser population is qualitatively similar
with Figure 1 in the publication (Kent2021).
There are quantitative differences however, for example, in Kent2021 after 10 years after image scan, the stroke-free
probability for the population with SBI and WMH was about 0.7 whereas in the simulated Kaiser population the same probability
was about 0.9.
It is interesting though that another feature is well reproduced in the simulation: the probability curves 
of SBI only and WMH only both in Kent2021 and the simulation are closer to each other compared to the no SBI or WMH 
or the SBI and WMH probability curves.

```{r}
setwd("/Users/deligkaris.1/OneDrive - The Ohio State University Wexner Medical Center/MICROSIM/NOTEBOOKS/DATA")
df <- read.csv("kaiserStrokeValidation.csv",header=TRUE)
survival_object <- Surv(time = df$time, event = df$event)
fit <- survfit(survival_object ~ scdGroup, data=df)
ggsurvplot(fit, 
           xlab="time [years]", ylab="stroke-free probability", 
           legend.labs=c("no SBI or WMH", "SBI only", "WMH only", "SBI and WMH"),
           legend=c(0.15, 0.5),
           break.time.by=1)$plot +
           ylim(c(0.75,1.))
```

## Dementia-free probability

Overall, the simulation reproduces qualitatively at least the results shown in Figure 2 in Kent2023.
Quantitatively however there are significant differences.

The MRI No group had the highest dementia-free probability in both the simulation and Kent2023.
The CT No group and MRI Mild group had probabilities that were close but not as close as Kent2023.
The MRI Moderate group had probabilities a bit higher than the MRI Unknown group, whereas the 
opposite was true in Kent2023.
The CT Unknown, CT Moderate and CT Severe had probabilities in the same order in the simulation and Kent2023.

Comparing the CT Severe probability 10 years after image scan, in Kent2023 this quantity is about 0.5 whereas in
the simulated Kaiser population it is about 0.87.


```{r}
setwd("/Users/deligkaris.1/OneDrive - The Ohio State University Wexner Medical Center/MICROSIM/NOTEBOOKS/DATA")
df <- read.csv("kaiserDementiaValidation.csv",header=TRUE)
blues <- brewer.pal(4, "Blues")
oranges <- brewer.pal(4, "Oranges")
greens <- brewer.pal(5, "Greens")
ct <- append(blues, values=greens[3], after=1)
mri <- append(oranges, values=greens[5], after=1)
palette <- c(ct, mri)
survival_object <- Surv(time = df$time, event = df$event)
fit <- survfit(survival_object ~ scdGroup, data=df)
fitLabels <- c("CT NO ", "CT UNKNOWN", "CT MILD", "CT MODERATE", "CT SEVERE",
                         "MRI NO ", "MRI UNKNOWN", "MRI MILD", "MRI MODERATE", "MRI SEVERE")
ggsurvplot(fit, 
           xlab="time [years]", ylab="dementia-free probability", 
           legend.labs=fitLabels,
           legend=c(0.15, 0.38),
           palette=palette,
           break.time.by=1,
           break.y.by=0.1,
           ylim=c(0.4,1))$plot 
```
```{r}
library(dplyr)
library(stringr)

fitdf <- data.frame(time = summary(fit)$time, surv = summary(fit)$surv, strata = summary(fit)$strata)
fitdf$severity <- as.character(fitdf$strata)
fitdf <- fitdf |> 
          mutate(
            severity = str_replace_all(severity, "scdGroup=0", "CTNo"),
            severity = str_replace_all(severity, "scdGroup=1", "CTNonspecified"),
            severity = str_replace_all(severity, "scdGroup=2", "CTMild"),
            severity = str_replace_all(severity, "scdGroup=3", "CTModerate"),
            severity = str_replace_all(severity, "scdGroup=4", "CTSevere"),
            severity = str_replace_all(severity, "scdGroup=5", "MRNo"),
            severity = str_replace_all(severity, "scdGroup=6", "MRNonspecified"),
            severity = str_replace_all(severity, "scdGroup=7", "MRMild"),
            severity = str_replace_all(severity, "scdGroup=8", "MRModerate"),
            severity = str_replace_all(severity, "scdGroup=9", "MRSevere"),
          ) |>
        mutate(probability = surv) |>
        select(time, probability, severity)
head(fitdf)
```




```{r}
library(tidyr)
library(dplyr)
setwd("/Users/deligkaris.1/OneDrive - The Ohio State University Wexner Medical Center/MICROSIM/NOTEBOOKS/DATA")
rdf <- read.csv("kaiserDementiaReference.csv",header=FALSE)
colNames <- rdf$V1
rdf <- data.frame(rdf |> select("V2":"V11") |> t())
names(rdf) <- colNames
rdf$x <- rdf$x / 12.

rdf <- rdf |> pivot_longer(cols=!x, names_to="severity", values_to="probability")
rdf

#fitLabels <- c("CTNo ", "CTNonspecified", "CTMild", "CTModerate", "CTSevere",
#                         "MRNo ", "MRNonspecified", "MRMild", "MRModerate", "MRSevere")
#fitLabels <- fitLabels[c(3, 4, 1, 2, 5, 8, 9, 6, 7, 10)]
#paletteN <- palette[c(3, 4, 1, 2, 5, 8, 9, 6, 7, 10)]

ggplot(data=rdf) +
  aes(x=factor(x), y=probability, color=severity) +
  geom_point() +
  scale_x_discrete() +
  scale_color_manual(values = paletteN) + #, labels=fitLabels) +
  theme_bw()

```

```{r}
setwd("/Users/deligkaris.1/OneDrive - The Ohio State University Wexner Medical Center/MICROSIM/NOTEBOOKS/DATA")
df <- read.csv("kaiserDementiaValidation.csv",header=TRUE)
blues <- brewer.pal(4, "Blues")
oranges <- brewer.pal(4, "Oranges")
greens <- brewer.pal(5, "Greens")
ct <- append(blues, values=greens[3], after=1)
mri <- append(oranges, values=greens[5], after=1)
palette <- c(ct, mri)
survival_object <- Surv(time = df$time, event = df$event)
fit <- survfit(survival_object ~ scdGroup, data=df)
fitLabels <- c("CT NO ", "CT UNKNOWN", "CT MILD", "CT MODERATE", "CT SEVERE",
                         "MRI NO ", "MRI UNKNOWN", "MRI MILD", "MRI MODERATE", "MRI SEVERE")
plot1 <- ggsurvplot(fit, 
           xlab="time [years]", ylab="dementia-free probability", 
           legend.labs=fitLabels,
           legend=c(0.15, 0.38),
           palette=palette,
           break.time.by=1,
           break.y.by=0.1,
           ylim=c(0.4,1))$plot 

rdf <- read.csv("kaiserDementiaReference.csv",header=FALSE)
colNames <- rdf$V1
rdf <- data.frame(rdf |> select("V2":"V11") |> t())
names(rdf) <- colNames
rdf$x <- rdf$x / 12.

rdf <- rdf |> pivot_longer(cols=!x, names_to="severity", values_to="probability")

plot1 +
  geom_point(data=rdf, aes(x=factor(x), y=probability, color=severity)) 
  #scale_x_discrete() +
  #scale_color_manual(values = paletteN) + #, labels=fitLabels) +
  #theme_bw()
```

