---
title: "pvdModelDevelopmentPart2"
output: html_document
date: "2024-01-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/deligkaris.1/OneDrive - The Ohio State University Wexner Medical Center/MICROSIM/NOTEBOOKS/DATA')
```

Load data and libraries

```{r, warning=FALSE}
library(lme4)
library(arm)
library(rstan)
library(rstanarm)

#with Rmarkdown notebooks this does not work outside this chunk
#dataDir<-"/Users/deligkaris.1/OneDrive - The Ohio State University Wexner Medical Center/MICROSIM/NOTEBOOKS/DATA"
#setwd(dataDir)
dataFile<-"pvdModelDevelopment.csv"
data<-read.csv(dataFile)
head(data)
```

Create the categorical variables.

```{r}
data$gender.f <- factor(data$gender)
data$smokingStatus0.f <- factor(data$smokingStatus0)
data$raceEthnicity0.f <- factor(data$raceEthnicity0)
```

Make sure all categorical variables are consistent with microsim

Male is 1, female is 2.

```{r}
str(data$gender.f)
```

Smoking status: never=0, former=1, current=2
Smoking status 0: never=0, former or current=1

```{r}
str(data$smokingStatus0.f)
```

raceEthnicity: mexican american=1, other hispanic=2, non hispanic white=3, non hispanic black=4, other=5
raceEthnicity0: mexican american or hispanic=1, non hispanic white=3, non hispanic black=4, other=5

```{r}
data$raceEthnicity0.f <- relevel(data$raceEthnicity0.f, ref="3")
str(data$raceEthnicity0.f)
```

The glm model.

```{r}
#model <- 
```

The stan glm model.

```{r}
model <- stan_glm(pvd1Binary ~ age0S + pulsePressure0S + tchdlRatio0 + gender.f + smokingStatus0.f + raceEthnicity0.f, refresh=0, data=data, family=binomial(link=logit))
print(summary(model), digits=2)
```

